Biology background

### 2015-04-12 Rmarkdown/biology_background.Rmd
Add my edits to Biology Background.

In this chapter I summarise the basic biology necessary for
understanding my thesis. I introduce DNA methylation, describe some of its biological roles and assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.

@@

The material in this chapter can be found in any good introductory molecular biology text, such as \citet{Alberts:2007vj}.

@@

The double helix is the most common, and most famous, structure of DNA. In the double helix, the two strands of DNA run in opposite directions to each other and are therefore _anti-parallel_. One strand is called the 5' strand (pronounced "5 prime strand") - also known as the sense strand, Crick strand or top strand - and the other strand called the 3' strand (pronounced "3 prime strand") - also known as the antisense strand, Watson strand or bottom strand. Along each strand of the double helix are the four DNA nucleobases (_bases_): adenine (_A_), cytosine (_C_), guanine (_G_) and thymine (_T_). These bases form _complementary_ base pairings, _A_ with _T_ and _C_ with _G_, along the DNA double helix. This is illustrated in Figure \ref{fig:double_helix}.

@@

DNA is able to self-replicate. This means that, for example, eukaryotic cells created during _mitosis_ contain the same DNA as the 'parent' cell. During DNA replication, the two strands are separated and each strand's _complementary_ DNA sequence is copied by an enzyme called _DNA polymerase_. It is because the two strands of DNA are complementary that ensures the daughter cell contains the same DNA sequence as the parent cell[^dna_replication].

@@

The core of a nucleosome consists of four pairs of histones, H2A, H2B, H3 and H4, which are consequently known as the _core histones_. A fifth histone, H1/H5, is known as the _linker histone_. Each of these histones has a 'tail' consisting of a string of amino acids. These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins \citep{Alberts:2007vj} and are discussed in Section \ref{sec:epigenetics} on epigenetics.

The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin. The linker sequences are between $20-60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.

Chromatin is frequently described as either _heterochromatin_ or _euchromatin_. Heterochromatin, or closed chromatin, is more tightly packed than euchromatin, or open chromatin. Heterochromatin is associated with transcriptionally repressed regions of the genome because the machinery required to translate DNA to RNA is unable to physically access the DNA. In contrast, euchromatin is associated with transcriptionally active regions of the genome since its more open nature allows easier access for the translational machinery.

The same region of an individual's genome may vary between heterochromatin and euchromatin states at different stages of the organism's life and in different cells of the organism. This is one mechanism which by gene expression is regulated.

@@

While eukaryotic genomes share the above-described features, and indeed share many regions of common DNA sequence, eukaryotic genomes come in many shapes and sizes. I only discuss the genomes of three organisms relevant to my thesis: Human (_Homo sapiens_), Mouse (_Mus musculus_) and Arabidopsis (_Arabidopsis thaliana_).

\subsubsection{Human genome}\label{sec:human_genome}

@@

\subsubsection{Mouse genome}\label{sec:mouse_genome}

@@

\subsubsection{Arabidopsis thaliana genome}\label{sec:arabidopsis_genome}

Arabidopsis is a small flowering plant. The most widely studied species is _A. thaliana_, which is diploid and has one of the smallest genomes known in the plant kingdom at 135 million base pairs long, which is packaged as five pairs of chromosomes ([http://plants.ensembl.org/Arabidopsis_thaliana/Info/Index](http://plants.ensembl.org/Arabidopsis_thaliana/Info/Index)). The small size of its genome, coupled with its small physical size, rapid development and its tolerance to both cross-pollination and self-pollination[^cross_and_self_pollination], make _A. thaliana_ an ideal organism for studying plant development.

[^cross_and_self_pollination]: The advantage in genetic studies of using a plant that is tolerant of self-pollination is that recessive mutations can be quickly made homozygous and thus expressed. Cross-pollination allows the creation of new strains with multiple mutations.

@@

[^indels]: An indel is a short insertion or deletion of DNA sequence in an organism, usually relative to a reference genome.

@@

_Epigenetics_, which I describe in the next section, plays a role in determining both the between-individual phenotypic variation as well as the within-individual phenotypic variation.

@@

_Epigenetics_ is a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, ... it means just what I choose it to mean - neither more nor less" \citep{carroll1897through}. I am sure to continue in this grand tradition.

@@

A popular contemporary definition of epigenetics is attributed to the epigeneticist Arthur Riggs - epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \citep[p1]{russo1996epigenetic}. The _epi_ prefix, derived from the Greek word for 'upon', 'near to', or 'in addition', emphasises the idea that epigenetics encodes information 'on top of' the DNA sequence. However, it is quite different to Waddington's original definition.

@@

Sir Adrian Bird, an esteemed British geneticist, and someone who is in favour of the broader definition, attempts to unite these definitions in the following \citep{Bird:2007em}:

> "The following could be a unifying definition of epigenetic events: the structural adaptation of chromosomal regions so as to register, signal or perpetuate altered activity states. This definition is inclusive of chromosomal marks [e.g. histone marks], because transient modifications associated with both DNA repair or cell-cycle phases and stable changes maintained across multiple cell generations qualify."

Regardless of which definition you subscribe to, an _epigenetic mark_ is the modification that causes this "epigenetic change". Actually, "causes" may be too strong a phrase, as much of current epigenetics research is in identifying associations rather than causations, and the question of whether the epigenetic mark is a cause or consequence of the ascribed function is oft-debated.

The _epigenome_ of a cell is the set of epigenetic marks present on the cell's genome. In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual. Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual as well will see in later chapters. _Epigenomics_ is the study of the epigenome, analogous to genomics being the study of the genome. However, one can rarely study the epigenome in isolation from the underlying genetic sequence, as there is evidence that the epigenetic variation is associated with genetic variation \citep{Zhang:2010bg, Bell:2011fm, vanEijk:2012hc, McVicker:2013ih}

One mark that most authors agree is an epigenetic mark is DNA methylation, which I describe in the next section and whose study is the focus of my thesis.

@@

DNA methylation is a chemical modification of DNA that can impart information on top of the DNA sequence. It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying]. It therefore fits into Arthur Riggs' aforementioned definition of an epigenetic modification \citep[p1]{russo1996epigenetic}. DNA methylation is found in bacteria, fungi, plants and animals.

@@

A major reason that DNA methylation is studied is that it is essential for normal development in mammals \citep{Li:1992bp}. It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression \citep{Razin:1991vj}, silencing of transposable elements\citep{Jones:2001eg}, X chromosome inactivation \citep{Mohandas:1981uf} and tumorigenesis \citep{Ehrlich:2002kz}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.3\textwidth]{../figures/5-methylcytosine.png}
\caption{Chemical structure of 5-methylcytosine. "5-methylcytosine". By Yikrazuul (Own work) [Public domain], via Wikimedia Commons [http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg](http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg)}
\label{fig:5mC}
\end{figure}

It is common to describe a cytosine as being 'methylated' or 'unmethylated', however, care must be taken in defining these terms. At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated. Double-stranded DNA, at least at _palindromic_ methylation loci[^palindromic_sequence], is generally symmetrically methylated, i.e., the loci on each strand are both methylated or both unmethylated. However, _hemimethylation_, where the methylation loci on one strand is methylated and its partner on the opposite strand is unmethylated, can and does occur.

[^palindromic_sequence]: A palindromic DNA sequence is one that is identical on when read in the 5' to 3' direction on both the original strand and the complementary strand of the double helix. For example, _CG_ is a palindromic sequence.

Within a diploid cell, a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other. While the former is more common, examples of the latter case, such as _allele-specific methylation_ \citep{Shoemaker:2010jz} and _genomic imprinting_ \citep{Li:1993ea}, are of considerable biological interest.

@@

CpGs are underrepresented in the human genome. The GC-content of each the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is between $41\%$. If the nucleotides were uniformly distributed across the human genome we would expect about $4.1\%$ of dinucleotides to be CpGs. Instead, only $1\%$ of dinucleotides are CpGs.

One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \citep{Scarano:1967ue}. Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinucleotide_frequency_hg19}). There are many other evolutionary pressures on the distribution of nucleotides in a genome. One effect of this is that the distribution of CpGs is far from uniform. In fact, CpGs tend to form clusters, which are termed CpG islands.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.3\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
\caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19). The expected frequency is computed under an 'independence' model, based on the observed frequencies of each nucleotide.}
\label{fig:dinucleotide_frequency_hg19}
\end{figure}

@@

One way to explore the distribution of CpGs in the human genome is to look at the distribution of distances from one CpG to the next, the _intra-pair distances_ (IPDs). Figure \ref{fig:IPD_ECDF_hg19} is a plot of the empirical cumulative distribution function of CpG IPDs for the human reference genome (hg19). We see that approximately $70\%$ of CpGs are within 100 bp of the next CpG. Figure \ref{fig:IPD_ECDF_hg19} also shows the expected IPD distribution under a model where CpGs are uniformly distributed along the genome with probability equal to the observed frequency of CpGs on each chromosome. By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more 'close' pairs and more 'distant' pairs than we would expect by chance.

@@

Figure \ref{fig:IPD_frequency_polygon_hg19} is an alternative way to visualise these data, which plots the percentage of pairs of CpGs with a given IPD. Figure \ref{fig:IPD_frequency_polygon_hg19} shows that there is a cluster of CpGs with $IPD < 10$. These largely correspond to CpGs that lie within what are called _CpG islands_ (CGIs).

@@

CpG islands contain the 20-40% of CpGs that are frequently unmethylated in mammalian genomes. CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}. The classical definition of a CpG island, given by \citep{GardinerGarden:1987we}, and used by, for example, the UCSC genome browser, is a region of the genome where:

@@

This definition was refined by \citet{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands. More recently, \citet{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region. I use this definition in my thesis.

@@

These 'sandy/beachy' metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons. CpG island shores, shelves and the open sea are all defined with respect to CpG islands. Specifically,

- CpG island shores are regions within 2 kb of CpG islands and have been demonstrated to have an increased variability of CpG methylation \citep{Irizarry:2008hg}. Shelves have a comparatively low density of CpGs.
- CpG island shelves are defined as regions within 2 kb of a CpG island shore \citep{Bibikova:2011fw}.
- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \citep{Sandoval:2011dj}.

@@

A subsequent paper from the same group extended this result. \citet{Lister:2011kg} reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines . Pluripotent cells includes embryonic stem cells along with induced pluripotent stem cell lines.

An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \citet{Lister:2013et}, where they found that neurons also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).

Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation. This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k beadchips, only measure CpG methylation. However, recent technological advances mean that cytosine methylation can be routinely assayed regardless of the sequence context. This discussed further in Section sec:dna_methylation_assays.

@@

Methylation is not the only chemical modification of cytosines, although it is by far the most common. These additional modifications may be thought of as 'modifications of a modification' since they extend 5-methylcytosine. Listed from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \citep{Plongthongkum:2014jg}. The biological significance of these marks is still being determined, in part because the assays for studying these are still in development and because their relatively scarcity in the genome means that experiments to detect these modifications are more difficult and expensive.

@@

\citet{Kriaucionis:2009bm} and \citet{Tahiliani:2009kl} discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC. This suggests a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks, as is discussed in the next section.

@@

A common analogy used in describing epigentic marks uses 'writers', 'readers' and 'erasers' \citep{Moore:2013in}. In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.

In mammalian cells, the writers are the DNA _methyltransferase_ (DNMT) enzymes. The DNMTs are commonly split into two groups, namely the maintenance methyltransferase and the _de novo_ methyltransferases. DNA methylation is not preserved by the DNA replication machinery and so it is the role of the maintainence methyltransferases to restore the methylation pattern on the daughter strand of DNA following DNA replication. In mammals, _DNMT1_ is known as the maintenance DNA methyltransferase. _DNMT3a_ and _DNMT3b_ are known as the _de nove_ methyltransferases, although these are also required for the maintenance of DNA methylation \citep{Jones:2009ji}. Both _DNMT1_ and _DNMT3b_ appear to be essential for mammalian development as mouse knockouts[^KO] for either gene is embryonically lethal \citep{Li:1992bp}. In contrast, mouse knockouts for _DNMT3a_ are runted but survive for ~4 weeks after birth \citep{Li:1992bp}.

@@

The removal or erasure of DNA methylation, called _demethylation_, is characterised as _passive_ loss or _active_ removal. Passive loss occurs when the maintenance methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division. This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions. This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of local tissue-specific differences in DNA methylation \citep{Irizarry:2009ky} nor the two stages of rapid global demethylation that occur during development \citep{Wu:2014gw}.

@@

1. The direct removal of the methyl group from _5mC_ is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
2. There is evidence that the DNA repair machinery can be co-opted to remove a methylated base or the surrounding region. The excised base or region is then repaired with unmethylated cytosines replacing 5mCs.
3. 5mC oxidation-dependent active DNA demethylation. This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ -> _5hmC_ -> _5fC_ -> _5caC_ reaction. The removal of _5hmC_, _5fC_ or _5caC_ is biochemically "easier" than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.

One question raised by point (3) is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks. This is an active area of investigation.

@@

Each of these pre-treatments can then be combined with a second step of amplification or hybridisation. For example,

@@

An exception to this classification scheme are a new class of assay that seek to directly 'read' whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA. For example, both \citet{Laszlo:2013kf} and \citet{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine is methylated.

Almost all assays of DNA methylation measure a population average from a pool of hundreds or thousands of cells. For a diploid organism, this is an average over several different levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay. Hundreds or thousands of cells are required in order to have sufficient material as input for the assay. Assays that require only a single cell as input do exist \citep[e.g.,]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.

@@

The throughput of an assay can be quantified in two ways. The first is per-sample throughput, which is how many measurements of DNA methylation are made per-sample[^throughput1]. This is typically what people mean when they describe an assay as being 'high-throughput' or 'low-throughput' and is the definition I use in the title of my thesis. Depending on your definition of 'high', a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample. The second definition of throughput is  related to cost, be it money or time, i.e., 'how many samples can I afford to analyse?''.

@@

In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays. In particular, I will describe in detail the 'gold standard; assay of DNA methylation, _whole-genome bisulfite-sequencing_ (WGBS), that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.

>>>>

In this chapter I summarise the basic biology necessary for
understanding my thesis. I introduce DNA methylation, describe some of its biological roles, and introduce assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.

@@

@@

The double helix is the most common, and most famous, structure of DNA. In the double helix, the two strands of DNA run in opposite directions to each other and are therefore _anti-parallel_. One strand is called the 5' strand (pronounced "5 prime strand" and also known as the sense strand, Crick strand or top strand) and the other strand called the 3' strand (pronounced "3 prime strand" and also known as the antisense strand, Watson strand or bottom strand). Along each strand of the double helix are the four DNA nucleobases (_bases_): adenine (_A_), cytosine (_C_), guanine (_G_) and thymine (_T_). These bases form _complementary_ base pairings, _A_ with _T_ and _C_ with _G_, along the DNA double helix. This is illustrated in Figure \ref{fig:double_helix}.

@@

DNA is able to self-replicate. This means that eukaryotic cells created during _mitosis_ contain the same DNA as the 'parent' cell. During DNA replication, the two strands are separated and each strand's _complementary_ DNA sequence is copied by an enzyme called _DNA polymerase_. It is because the two strands of DNA are complementary that ensures the daughter cell contains the same DNA sequence as the parent cell[^dna_replication].

@@

The core of a nucleosome consists of four pairs of histones, H2A, H2B, H3 and H4, which are consequently known as the _core histones_. A fifth histone, H1/H5, is known as the _linker histone_. Each of these histones has a 'tail' consisting of a string of amino acids. These tails can undergo _post-translational modifications_, such as methylation, acetylation and phosphorylation, which can alter their interactions with DNA and nuclear proteins \citep{Alberts:2007vj}. Histone modifications are discussed in Section \ref{sec:epigenetics}.

The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin. The linker sequences are between $20--60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.

Chromatin is frequently described as either 'closed' or 'open'. Closed chromatin, _heterochromatin_, is more tightly packed than the open _euchromatin_. Heterochromatin is associated with transcriptionally repressed regions of the genome because the machinery required to translate DNA to RNA is unable to physically access the DNA. In contrast, euchromatin is associated with transcriptionally active regions of the genome since its more open nature allows easier access for the translational machinery.

The same region of an individual's genome may vary between heterochromatin and euchromatin states at different stages of the organism's life and in different cells of the organism. This is one mechanism by which gene expression is regulated.

@@

While eukaryotic genomes share the above-described features, and indeed share many regions of common DNA sequence, eukaryotic genomes come in many shapes and sizes. I only discuss the genomes of two species relevant to my thesis: _Homo sapiens_ (human) and _Mus musculus_ (house mouse).

\subsubsection{The human genome}\label{sec:human_genome}

@@

\subsubsection{The mouse genome}\label{sec:mouse_genome}

@@

@@

[^indels]: An example of a larger genetic variant is an _indel_, which is a short insertion or deletion of sequence in an organism's DNA, usually with respect to a reference genome.

@@

_Epigenetics_, which I describe in the next section, plays a role in determining both the between-individual phenotypic variation, as well as the within-individual phenotypic variation. DNA methylation, the focus of my thesis, is the prototypical and most well-studied epigenetic modification.

@@

Interest in epigenetics has grown remarkably in recent years. However, _epigenetics_ is also a real Humpty Dumpty phrase; each author seems to believe that, "When I use [the] word, \dots it means just what I choose it to mean --- neither more nor less" \citep{carroll1897through}. As noted by \citet{Deans:2015kt}, "the unfortunate fact is that the increased use of the term _epigenetics_ is likely due more to inconsistencies in its definition than to a consensus of interest among scientists".

@@

A popular contemporary definition of epigenetics is attributed to the epigeneticist Arthur Riggs --- epigenetics is "the study of mitotically and/or meiotically heritable changes in gene function that cannot be explained by changes in DNA sequence[^mitotic_meiotic]" \citep[pp. 1]{russo1996epigenetic}. The _epi_ prefix, derived from the Greek word for 'upon', 'near to', or 'in addition', emphasises the idea that epigenetics encodes information 'on top of' the DNA sequence. However, it is quite different to Waddington's original definition.

@@

Sir Adrian Bird, an esteemed British geneticist, attempts to unite these definitions \citep{Bird:2007em}:

> "The following could be a unifying definition of epigenetic events: the structural adaptation of chromosomal regions so as to register, signal or perpetuate altered activity states. This definition is inclusive of chromosomal marks [e.g., histone modifications], because transient modifications associated with both DNA repair or cell-cycle phases and stable changes maintained across multiple cell generations qualify."

Regardless of which definition you subscribe to, an _epigenetic mark_ is the modification that causes this 'epigenetic change'. In fact, 'causes' may be too strong a claim, as much of current epigenetics research is in identifying associations rather than causations, and the question of whether the epigenetic mark is a cause or consequence of the ascribed function is oft-debated.

The _epigenome_ of a cell is the set of epigenetic marks present on the cell's genome. In contrast to the genome, which is identical between cells within an individual, the epigenome is highly variable between cells within an individual. Indeed, we can identify variation for a single epigenetic mark within cells of the same cell type from the same individual (see Chapter \ref{chap:co-methylation}). _Epigenomics_ is the study of the epigenome, analogous to genomics being the study of the genome. However, one can rarely study the epigenome in isolation from the underlying genetic sequence, as there is evidence that the epigenetic variation is associated with genetic variation \citep{Zhang:2010bg, Bell:2011fm, vanEijk:2012hc, McVicker:2013ih}

One mark that most authors agree is an epigenetic mark is DNA methylation, which I describe in the next section and the study of which is the focus of my thesis.

@@

DNA methylation is a chemical modification of DNA that can impart information on top of the DNA sequence. It is heritable during mitotic cell division, which means that it is faithfully copied across to the daughter cell during cell division[^dnam_copying]. It therefore fits into Arthur Riggs' aforementioned definition of an epigenetic modification. DNA methylation is found in bacteria, fungi, plants and animals.

@@

A major reason that DNA methylation is studied is that it is essential for normal development in mammals \citep{Li:1992bp}. It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression \citep{Razin:1991vj}, silencing of transposable elements \citep{Jones:2001eg}, X chromosome inactivation \citep{Mohandas:1981uf} and tumorigenesis \citep{Ehrlich:2002kz}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.3\textwidth]{../figures/5-Methylcytosine.pdf}
\caption{Chemical structure of 5-methylcytosine. "5-methylcytosine". By Yikrazuul (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg}}
\label{fig:5mC}
\end{figure}

It is common to describe a cytosine as being 'methylated' or 'unmethylated', however, care must be taken when using these terms. At the lowest level, the level of single-stranded DNA, methylation is a binary event: a cytosine is either methylated or unmethylated. Double-stranded DNA, at least at _palindromic_ methylation loci[^palindromic_sequence], is generally symmetrically methylated, i.e., the loci on each strand are both methylated or both unmethylated. However, _hemimethylation_, where the methylation loci on one strand is methylated and its partner on the opposite strand is unmethylated, can and does occur.

[^palindromic_sequence]: A palindromic DNA sequence is one that is identical when read in the 5' to 3' direction on both the original strand and the complementary strand of the double helix. For example, _CG_ is a palindromic sequence.

Within a diploid cell, a particular cytosine may be unmethylated or methylated on both homologous chromosomes or methylated on one chromosome and unmethylated on the other. While the former is more common, examples of the latter case, such as _allele-specific methylation_ \citep{Shoemaker:2010jz} and _genomic imprinting_ \citep{Li:1993ea}, are important epigenetic phenomena.

@@

CpGs are underrepresented in the human genome. The _GC-content_ of the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is approximately $41\%$. If these nucleotides were uniformly distributed across the genome then we would expect about $4.1\%$ of _dinucleotides_, pairs of nucleotides, to be CpGs. Instead, only $1\%$ of dinucleotides are CpGs.

One reason for the relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \citep{Scarano:1967ue}. Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinucleotide_frequency_hg19}). There are many other evolutionary pressures on the distribution of nucleotides in a genome. One effect of this is that the distribution of CpGs is far from uniform. In fact, CpGs tend to form clusters, which are termed CpG islands.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
\caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19). The expected frequency is computed under an 'independence' model, based on the observed frequencies of each nucleotide.}
\label{fig:dinucleotide_frequency_hg19}
\end{figure}

@@

One way to explore the distribution of CpGs in the human genome is to look at the distribution of distances from one CpG to the next, the _intra-pair distances_ (IPDs). Figure \ref{fig:IPD_ECDF_hg19} is a plot of the empirical cumulative distribution function of CpG IPDs for the human reference genome (hg19). We see that approximately $70\%$ of CpGs are within 100 bp of the next CpG. Figure \ref{fig:IPD_ECDF_hg19} also shows the expected IPD distribution under a model where CpGs are uniformly distributed along the genome with probability equal to the observed frequency of CpGs on each chromosome. By comparing the observed IPDs to the expected IPDs we see that the distribution of distances between CpGs has more 'close' pairs than we would expect by chance.

@@

Figure \ref{fig:IPD_frequency_polygon_hg19} is an alternative way to visualise these data by plotting the percentage of pairs of CpGs with a given IPD. Figure \ref{fig:IPD_frequency_polygon_hg19} shows that there is a cluster of CpGs with $IPD < 10$. These largely correspond to CpGs that lie within what are called _CpG islands_ (CGIs).

@@

CpG islands contain the 20-40% of CpGs that are frequently unmethylated in mammalian genomes. CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}. The classical definition of a CpG island, given by \citet{GardinerGarden:1987we}, and used by the popular UCSC genome browser ([http://genome.ucsc.edu/cgi-bin/hgGateway](http://genome.ucsc.edu/cgi-bin/hgGateway)), is a region of the genome where:

@@

This definition was refined by \citet{Takai:2002jb} to exclude _Alu_-repetitive elements, which are otherwise misclassified as _bona fide_ CpG islands. More recently, \citet{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region. In my thesis, I use the definition given by \citet{Wu:2010do}.

@@

These 'sandy/beachy' metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons. CpG island shores, shelves and the open sea are all defined with respect to CpG islands:

- CpG island shores are regions within 2 kb of CpG islands. These have been demonstrated to have an increased variability of CpG methylation \citep{Irizarry:2009ky}.
- CpG island shelves are defined as regions within 2 kb of a CpG island shore \citep{Bibikova:2011fw}.
- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \citep{Sandoval:2011dj}.

@@

A subsequent paper from the same group extended this result. \citet{Lister:2011kg} reported that, more generally, non-CpG cytosines account for $20-30\%$ of cytosines with statistically significant evidence of methylation in _pluripotent_ cell lines . Pluripotent cells includes embryonic stem (_ES_) cells and induced pluripotent stem (_iPS_) cell lines.

An exception to the rule that non-CpG methylation is largely restricted to pluripotent cells is provided by \citet{Lister:2013et}, who found that neurons also have non-CpG methylation, albeit at a lower level ($1.3-1.5\%$ of all non-CpG cytosines were methylated).

Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation. This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k microarrays, only measure CpG methylation. However, recent technological advances mean that cytosine methylation can be routinely assayed regardless of the sequence context (see Section \ref{sec:dna_methylation_assays}).

@@

Methylation is not the only chemical modification of cytosines, although it is by far the most common. Listed from most frequent to least frequent, these are are 5-hydroxymethylcytosine (_5hmC_), 5-formylcytosine (_5fC_) and 5-carboxylcytosine (_5caC_) \citep{Plongthongkum:2014jg}. The biological significance of these marks is still being determined, in part because the assays for studying these are still in development and because their relatively scarcity in the genome means that experiments to detect these modifications are more difficult and expensive.

@@

\citet{Kriaucionis:2009bm} and \citet{Tahiliani:2009kl} discovered that the TET enzymes can convert 5mC to 5hmC, 5hmC to 5fC and 5fC to hcaC. This suggests a role for 5hmC, 5fC and 5caC in the process of removing 5mC marks.

@@

A frequently used analogy when describing epigentic marks refers to 'writers', 'readers' and 'erasers' \citep[e.g.,]{Moore:2013in}. In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.

In mammalian cells, the writers are the DNA _methyltransferase_ (DNMT) enzymes. The DNMTs are commonly split into two groups, namely the maintenance methyltransferases and the _de novo_ methyltransferases.

DNA methylation is not preserved by the DNA replication machinery and so it is the role of the maintainence methyltransferases to restore the methylation pattern on the daughter strand of DNA following DNA replication. In mammals, _DNMT1_ is known as the maintenance DNA methyltransferase.

_DNMT3a_ and _DNMT3b_ are known as the _de novo_ methyltransferases, although these are also required for the maintenance of DNA methylation \citep{Jones:2009ji}. Both _DNMT1_ and _DNMT3b_ appear to be essential for mammalian development since mouse knockouts[^KO] for either gene are embryonically lethal \citep{Li:1992bp}. In contrast, mouse knockouts for _DNMT3a_ are runted but survive for ~4 weeks after birth \citep{Li:1992bp}.

@@

The removal or erasure of DNA methylation, called _demethylation_, may be characterised as _passive_ loss or _active_ removal. Passive loss occurs when the maintenance methyltransferases do not efficiently perform their role of restoring DNA methylation following cell division. This leads to a gradual, stochastic and genome-wide loss of DNA methylation after multiple cell divisions. This form of passive demethylation, sometimes called replication-dependent demethylation, cannot explain observations of local tissue-specific differences in DNA methylation \citep{Irizarry:2009ky} nor the two stages of rapid global demethylation that occur during development \citep{Wu:2014gw}.

@@

1. The direct removal of the methyl group from _5mC_ is considered unlikely due to the strong carbon-carbon bond between the methyl group and the cytosine.
2. There is evidence that the DNA repair machinery can be co-opted to remove a methylated base or the surrounding region. The excised base or region is then repaired with unmethylated cytosines replacing 5mCs.
3. 5mC oxidation-dependent active DNA demethylation. This follows from the observation that the TET enzymes can iteratively oxidate a _5mC_ $\to$ _5hmC_ $\to$ _5fC_ $\to$ _5caC_ reaction. The removal of _5hmC_, _5fC_ or _5caC_ is biochemically 'easier' than the removal of _5mC_ and could occur via a more efficient form of replication-dependent demethylation, direct removal of the oxidized methyl group or through the DNA repair machinery.

One question raised by point (3) is whether _5hmC_, _5fC_ and _5caC_ are simply intermediate products in an active demethylation cycle or if they themselves are bona fide epigenetic marks. This is an active area of research.

@@

Following treatment, DNA methylation can be assayed using standard techniques such as:

@@

An exception to this classification scheme are a new class of assays that seek to directly 'read' whether a position is methylated or unmethylated without requiring a pre-treatment of the DNA. For example, both \citet{Laszlo:2013kf} and \citet{Schreiber:2013in} measure the change in current as a DNA molecule passes through a nanopore to infer whether a cytosine is methylated.

Almost all assays of DNA methylation measure a population average from a pool of hundreds or thousands of cells. For a diploid organism, this is an average over several distinct levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay. Hundreds or thousands of cells are required in order to have sufficient material as input for the assay. Assays that require only a single cell as input do exist \citep[e.g.,]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.

@@

The throughput of an assay can be quantified in two ways. The first is per-sample throughput, which is how many measurements of DNA methylation are made per-sample[^throughput1]. This is typically what people mean when they describe an assay as being 'high-throughput' or 'low-throughput' and is the definition I use in the title of my thesis. Depending on your definition of 'high', a high-througput assay will produce on the order of tens of thousands to billions of measurements per sample. The second definition of throughput is  related to cost, be it money or time, i.e., 'how many samples can I afford to analyse?'.

@@

In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays. In particular, I will describe in detail the 'gold standard' assay of DNA methylation, _whole-genome bisulfite-sequencing_, that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.
